---
title: "Homework 2"
author: "Ruichen Yang"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```


## Problem 1

```{r}
subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE
) |>
  clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entrance_type,
    entry,
    vending,
    ada
  ) |>
  mutate(
  entry = entry == "YES"
)
```


The cleaned subway dataset contains `r nrow(subway_df)` observations and `r ncol(subway_df)` variables. It includes station name and line, station latitude and longitude, routes served, entrance type, entry permission, vending status, and ADA compliance. I cleaned the data by standardizing variable names, selecting the requested variables, and converting the entry variable from YES/NO to a logical variable. These data are not fully tidy because the routes served are stored across multiple columns (`route1` through `route11`) rather than in a single route variable.

```{r}
station_count = subway_df |>
  distinct(line, station_name) |>
  nrow()

station_count
```
```{r}
ada_station_count = subway_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

ada_station_count
```

```{r}
entry_prop_no_vending = subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry)
  ) |>
  pull(proportion)

entry_prop_no_vending
```

```{r}
subway_long = subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_transform = list(route_name = as.character),
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )

head(subway_long)
```

```{r}
a_train_station_count = subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_station_count
```

```{r}
a_train_ada_count = subway_long |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name) |>
  nrow()

a_train_ada_count
```

There are `r station_count` distinct subway stations in the dataset, where stations are identified by both line and station name. Of these, `r ada_station_count` stations are ADA compliant. Among station entrances and exits without vending, approximately `r round(entry_prop_no_vending * 100, 1)`% allow entrance.

After reorganizing the route variables into a tidy format, there are `r a_train_station_count` distinct stations that serve the A train. Of these stations, `r a_train_ada_count` are ADA compliant.