---
title: "Homework 2"
author: "Ruichen Yang"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```


## Problem 1

```{r}
subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE
) |>
  clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entrance_type,
    entry,
    vending,
    ada
  ) |>
  mutate(
  entry = entry == "YES"
)
```


The cleaned subway dataset contains `r nrow(subway_df)` observations and `r ncol(subway_df)` variables. It includes station name and line, station latitude and longitude, routes served, entrance type, entry permission, vending status, and ADA compliance. I cleaned the data by standardizing variable names, selecting the requested variables, and converting the entry variable from YES/NO to a logical variable. These data are not fully tidy because the routes served are stored across multiple columns (`route1` through `route11`) rather than in a single route variable.

```{r}
station_count = subway_df |>
  distinct(line, station_name) |>
  nrow()

station_count
```
```{r}
ada_station_count = subway_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

ada_station_count
```

```{r}
entry_prop_no_vending = subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry)
  ) |>
  pull(proportion)

entry_prop_no_vending
```

```{r}
subway_long = subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_transform = list(route_name = as.character),
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )

head(subway_long)
```

```{r}
a_train_station_count = subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_station_count
```

```{r}
a_train_ada_count = subway_long |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name) |>
  nrow()

a_train_ada_count
```

There are `r station_count` distinct subway stations in the dataset, where stations are identified by both line and station name. Of these, `r ada_station_count` stations are ADA compliant. Among station entrances and exits without vending, approximately `r round(entry_prop_no_vending * 100, 1)`% allow entrance.

After reorganizing the route variables into a tidy format, there are `r a_train_station_count` distinct stations that serve the A train. Of these stations, `r a_train_ada_count` are ADA compliant.


## Problem 2

```{r}
excel_sheets("data/202610 Trash Wheel Collection Data.xlsx")
```

```{r}
mr_trash = read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N743"
) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

dim(mr_trash)
```

```{r}
professor_trash = read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel",
  range = "A2:M140"
) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    plastic_bottles = na_if(as.character(plastic_bottles), "dive"),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )

dim(professor_trash)
```

```{r}
gwynnda_trash = read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynnda the Good Wheel of the W",
  range = cell_limits(c(2, 1), c(NA, 12))
) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Gwynnda"
  )

dim(gwynnda_trash)
```

```{r}
trash_wheel_df = bind_rows(
  mr_trash |> mutate(year = as.integer(year)),
  professor_trash |> mutate(year = as.integer(year)),
  gwynnda_trash |> mutate(year = as.integer(year))
)

dim(trash_wheel_df)
```

```{r}
professor_total_weight = trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  ) |>
  pull(total_weight)

professor_total_weight
```


```{r}
gwynnda_cigarette_butts = trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_butts = sum(cigarette_butts, na.rm = TRUE)
  ) |>
  pull(total_butts)

gwynnda_cigarette_butts
```

The combined Trash Wheel dataset contains `r nrow(trash_wheel_df)` observations and `r ncol(trash_wheel_df)` variables. The data include information on the date of each dumpster collection, weight and volume of trash collected, counts of different types of litter such as plastic bottles, polystyrene, cigarette butts, glass bottles, plastic bags, wrappers, and sports balls, as well as the Trash Wheel responsible for each collection. I imported and cleaned the three Trash Wheel datasets separately, removed non-data rows, standardized variable names and types, and then combined them into a single tidy dataset.

For the available data, Professor Trash Wheel collected a total of `r professor_total_weight` tons of trash. Gwynnda collected `r format(gwynnda_cigarette_butts, big.mark = ",", scientific = FALSE)` cigarette butts in June 2022.


## Problem 3


```{r}
mci_baseline = read_csv(
  "data/MCI_baseline.csv",
  skip = 1,
  na = "."
) |>
  clean_names() |>
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  ) |>
  filter(
    is.na(age_at_onset) | age_at_onset > current_age
  )

dim(mci_baseline)
```
```{r}
baseline_summary = mci_baseline |>
  summarize(
    participants = n(),
    developed_mci = sum(!is.na(age_at_onset)),
    mean_baseline_age = mean(current_age),
    prop_women_apoe4 = mean(apoe4[sex == "Female"] == "Carrier")
  )

baseline_summary
```


The baseline dataset initially contained 483 participants. I skipped the non-data first row during import, treated "." as missing values, cleaned the variable names, and encoded sex and APOE4 carrier status as categorical variables rather than numeric variables. I then removed participants who already had MCI at baseline, leaving `r nrow(mci_baseline)` eligible participants.

Among these participants, `r baseline_summary$developed_mci` developed MCI during follow-up. The average baseline age was `r round(baseline_summary$mean_baseline_age, 2)` years. Among women in the study, `r round(baseline_summary$prop_women_apoe4 * 100, 1)`% were APOE4 carriers.


```{r}
amyloid_raw = read_csv(
  "data/mci_amyloid.csv",
  skip = 1,
  na = c(".", "NA", "Na"),
  show_col_types = FALSE
) |>
  clean_names()
```


```{r}
amyloid_long = amyloid_raw |>
  pivot_longer(
    cols = baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio"
  ) |>
  mutate(
    time_years = recode(
      visit,
      "baseline" = 0,
      "time_2" = 2,
      "time_4" = 4,
      "time_6" = 6,
      "time_8" = 8
    ),
    amyloid_ratio = as.numeric(amyloid_ratio)
  )

dim(amyloid_long)
head(amyloid_long)
```


```{r}
baseline_only = mci_baseline |>
  select(id) |>
  anti_join(
    amyloid_raw |> select(study_id),
    by = c("id" = "study_id")
  )

amyloid_only = amyloid_raw |>
  select(study_id) |>
  anti_join(
    mci_baseline |> select(id),
    by = c("study_id" = "id")
  )

baseline_only
amyloid_only
```
```{r}
mci_combined = mci_baseline |>
  inner_join(
    amyloid_long,
    by = c("id" = "study_id")
  ) |>
  arrange(id, time_years)

dim(mci_combined)

mci_combined |>
  distinct(id) |>
  nrow()
write_csv(
  mci_combined,
  "data/mci_combined.csv"
)
```


The amyloid dataset initially contained 487 participants, with amyloid biomarker measurements recorded at baseline and at approximately 2, 4, 6, and 8 years of follow-up. I skipped the non-data first row during import, treated the dataset's missing-value codes as missing values, cleaned the variable names, and reshaped the repeated biomarker measurements from wide to long format. The resulting long-format dataset contains one row for each participant-visit combination, with separate variables for the visit, time since baseline, and amyloid ratio.

There were 8 participants who appeared only in the baseline dataset and 16 participants who appeared only in the amyloid dataset. After retaining only participants present in both datasets, the combined dataset contains 471 unique participants, 2,355 observations, and 9 variables. The combined dataset includes baseline demographic and MCI information together with longitudinal amyloid measurements. The final dataset was exported as `data/mci_combined.csv`.


